Manuscript

### 2025-03-18 Manuscript_MD_ScoreReLiability.qmd
change & shorten first discussion

$$\tau^2[\delta_0] \approx \frac{\tau^2[MD]}{\mu[\sigma_x]^2} + \frac{\mu[MD]^2}{\mu[\sigma_x]^4} var[SD]$$ {#eq-ratio_d0}

@@

Instead, what is guaranteed is that the expected value of $\sigma_T$ is
smaller than the expected value of $\sigma_X$. Similarly, the
heterogeneity in $\sigma_T$ is guaranteed to be smaller than the
uncorrected heterogeneity in $\sigma_X$. However, since the expected
value of either standard deviation is placed in the denominator of the
function and the variance of either is placed in the numerator of the
function, @eq-hs is not sufficient to explain how heterogeneity in ES
changes due to the attenuation correction. Instead, we introduce two
additional metrics.

@@

$R_1$ describes the relative size of mean true score variance, compared
to the mean observed score variance. This informs us, how much of the
mean total score variance can be attributed to mean true score variance.
Therefore, this metric is close to the average score reliability. Even
though formally $R_1$ and the meta-analytic mean score reliability
$\mu_{\rho_{xx'}}$ are not the same value, in practice we observe little
to no differences between the two parameters. Large values indicate that
correcting the individual score variances (or standard deviations for
that matter) using the attenuation correction procedure should lead to
very little change, as most of the variation observed were due to
genuine differences in true scores. Small values indicate the opposite,
the observed score variance was due to a larger amount of random error
score variance, leading to a stronger correction.

@@

Both metrics $R_1$ and $R_2$ make use of parameters of the distribution
of true and total score variance, not standard deviations. This choice
was made to facilitate the role of score reliability in the subsequent
discussion, as score reliability itself is defined by variances, not
standard deviations (see @eq-rel).

Using the metrics $R_1$ and $R_2$, defined in [Equations @eq-R1] and
[-@eq-R2], we attempt to understand under which circumstances the
inequality defined in @eq-ineq_init holds true. However, while the
metrics make use of mean and heterogeneity of the score variances,
@eq-ineq_init makes use of mean and heterogeneity of the standard
deviations. In order to incorporate the metrics, it is necessary to
reparameterise the equation. Using the delta method to approximate mean
value and heterogeneity of variance from those of the standard
deviations, we know that (ref):
$$\mu^2_{\sigma_X} \approx \mu_{\sigma^2_X} \quad \quad  and \quad \quad 
\tau^2_{\sigma_X} \approx \frac{\tau^2_{\sigma_X^2}}{4\mu_{\sigma^2_X}}$$ {#eq-delta_method}

While @eq-delta_method only explicitly contains parameters for the
observed score variance/standard deviation, the same can be done using
the true score variance/standard deviation. Using the approximates
defined in @eq-delta_method for observed and true score variance, we
arrive at a new inequality in @eq-ineq_CV.

$$\frac{\tau_{MD}^2}{\mu_{\sigma^2_t}} + \frac{\mu_{MD}^2}{\mu^2_{\sigma_t^2}} \frac{\tau^2_{\sigma_t^2}}{4\mu_{\sigma^2_t}} > \frac{\tau_{MD}^2}{\mu_{\sigma^2_x}} + \frac{\mu_{MD}^2}{\mu^2_{\sigma_x^2}} \frac{\tau^2_{\sigma_x^2}}{4\mu_{\sigma^2_x}}$$ {#eq-ineq_CV}

@eq-ineq_CV describes that the heterogeneity in corrected ES is larger
than the heterogeneity in uncorrected ES, using parameters of the
distributions of mean difference MD, true score variance $\sigma^2_T$
and observed score variance $\sigma^2_X$. By introducing the metrics
$R_1$ and $R_2$ to this inequality, we can begin to disentangle which
conditions need to be fulfilled for the inequality to hold. Rearranging
the terms in [Equations @eq-R1] and [-@eq-R2], and subsequently entering
these terms into Equation 18 leads to the following
$$\frac{\tau_{MD}^2}{R_1 \mu_{\sigma^2_x}} + \frac{\mu_{MD}^2}{R_1^2 \mu^2_{\sigma^2_x}} \frac{R_2 \tau^2_{\sigma_x^2}}{4 R_1 \mu_{\sigma^2_x}} > \frac{\tau_{MD}^2}{\mu_{\sigma^2_x}} + \frac{\mu_{MD}^2}{\mu^2_{\sigma_x^2}} \frac{\tau^2_{\sigma_x^2}}{4\mu_{\sigma^2_x}}$$ {#eq-ineq_R}

@eq-ineq_R, again, describes the inequality that heterogeneity in
corrected ES is larger than the heterogeneity in uncorrected ES,
incorporating the metrics $R_1$ and $R_2$. This equation alone is not
sufficient to identify the relevant circumstances required for the
inequality to hold. However, rearranging the terms from @eq-ineq_R leads
to @eq-ineq_fin:

$$ \tau_{MD}^2\left(\frac{1}{R_1} - 1\right) + \mu_{MD}^2 \frac{1}{4} CV^2_{\sigma^2_X} \left(\frac{R_2}{R_1^3}  - 1\right) > 0$$ {#eq-ineq_fin}

The circumstances under which the inequality described in Equation 20
holds, are circumstances where the claims made in W&D and H&S are
directly contradicted, as in those cases the ES heterogeneity is larger
after applying an attenuation correction procedure. Concerning the terms
in @eq-ineq_fin, we know that all terms left-hand of the inequality,
outside of the brackets ($\tau^2_{MD}$, $\mu^2_{MD}$ and
$CV^2_{\sigma^2_X}$), are bound to be positive. For the inequality to
hold, we need the left-hand side of the equation to remain positive,
larger than zero. We can distill two scenarios, under which this
equation should hold: (a) one of the terms inside the brackets
($\frac{1}{R_1} - 1$ and $\frac{R^2_2}{R_1}  - 1$) is positive and large
enough, so that the second term not containing that same bracket is
positively dominated by the first term; or (b) both terms inside the
brackets need to be positive.

Generally, we know that both $R_1$ and $R_2$ are bound to be positive,
as both contain different, strictly positive parameters of the
distributions of true and observed score variance. Additionally, we know
that $R_1$ is equivalent to the average score reliability, and therefore
bound between $0\le R_1\le1$. Therefore, the term inside the first
bracket is bound to be positive ($\frac{1}{R_1} – 1$). Similarly, $R_2$,
as the ratio of true and error score variance heterogeneity, is bound
between $0\le R_2\le1$. If $R_1^3$ is smaller than $R_2$, then the term
inside the second bracket turns negative. Generally, this means that the
inequality can only be violated if the reduction in relative
heterogeneity of score variance ($R_2$) is larger than the the average
score reliability to the power of 3.

Concerning the two scenarios, for scenario (a) to hold true, since the
term involving only $R_1$ in the first bracket can only be positive, the
second term involving both metrics would need to be negative. For the
first term to positively dominate the second term, however, we would
need to make additional assumptions about the size of the parameters
outside of the brackets, specifically about $\tau^2_{MD}$ and
$\mu_{MD}$. This is something we would prefer to avoid, as it involves
additional assumptions about the individual raw effect sizes of the
different phenomena. Additionally, it can be said that as long as the
term in the second bracket involving both metrics is positive, scenario
(b) is true by default. This occurs any time $R_1^3 > R_2$. For example,
assuming the average score reliability is about .8, if $R_2$ is at least
.512 or larger, the inequality holds true and ES heterogeneity grows
larger as a consequence of the attenuation correction procedure,
contrary to claims made in W&D and H&S. This means that for the
exemplary score reliability of .8, a reduction in relative heterogeneity
in score variance of 48.8% ($(1 – R_2)100\%$) or more is necessary for
scenario (b) to no longer be true. The smaller the average score
reliability, the smaller the $R_2$-metric may be for the inequality to
still hold true.

Keeping this in mind, even if $R_2$ is smaller than .512, it does not
immediately imply that ES heterogeneity is smaller in $\delta$. For
example, it is unlikely that an $R_2 = .5$ leads to a term that is not
positively dominated by the first term involving only $R_1$. However,
where exactly the cut-off is, where ES heterogeneity starts shrinking
due to the attenuation correction procedure, can not be easily defined
without additional assumptions about the distribution of $MD$.

Generally speaking, the attenuation correction leads to a reduction both
in mean value of $\sigma^2_X$ and in its heterogeneity. While a
reduction in the former typically leads to an increase in heterogeneity
of $d_c$, the latter has the potential to compensate for that increase
and thereby reduce that heterogeneity after the attenuation correction.
This reduction in ES heterogeneity can only occur, if the reduction in
score variance heterogeneity is substantially stronger than the
reduction in mean score variance. An extreme version of such a situation
might occur, if the true score variance is identical across all samples.
This would imply that all heterogeneity in score variance identified can
be attributed to differences in how much random noise is found in the
scores across the samples. In such a case, $R_2$ would be zero and a
reduction in ES heterogeneity due to the attenuation correction would be
guaranteed. This also implies that the lower $R_2$ and the larger the
average score reliability $R_1$, the more likely it is that the
attenuation correction actually does lead to a reduction in ES
heterogeneity.

>>>>

$$\tau^2[\delta_0] \approx \frac{\tau^2[MD]}{\mu[\sigma_x]^2} + \frac{\mu[MD]^2}{\mu[\sigma_x]^4} var[\sigma_x]$$ {#eq-ratio_d0}

@@

Instead, what is guaranteed is that the expected value of $\sigma_T$ is
smaller than the expected value of $\sigma_X$. Similarly, the
heterogeneity in $\sigma_T$ is guaranteed to be smaller than the
uncorrected heterogeneity in $\sigma_X$. However, since the expected
value of either standard deviation is placed in the denominator of the
function and the variance of either is placed in the numerator of the
function, @eq-hs is not sufficient to explain how heterogeneity in ES
changes due to the attenuation correction.

As long as the inequality defined in @eq-ineq_init holds true, an
attenuation correction procedure would lead to larger estimates of ES
heterogeneity. Conditions under which @eq-ineq_init does not hold are
conditions under which the claims made in W&D and H&S are met. However,
the inequality in this form does not convey sufficient information about
the conditions under which it can potentially hold. In Appendix A, we
rearranged @eq-ineq_init, so we end up with a form that we can use to
actually derive conditions under which ES heterogeneity can be inflated
or deflated by differences in score reliability.

To do so, we introduce two additional metrics.

@@

$R_1$ describes the relative size of mean true score variance, compared
to the mean observed score variance. This informs us, how much of the
mean total score variance can be attributed to mean true score variance.
Therefore, this metric is close to the average score reliability. Even
though formally $R_1$ and the meta-analytic mean score reliability
$\mu_{\rho_{xx'}}$ do not share not the same value, in practice we
observe little to no differences between the two parameters. Large
values indicate that correcting the individual score variances (or
standard deviations for that matter) using the attenuation correction
procedure should lead to very little change in ES, as most of the
variation observed were due to genuine differences in true scores. Small
values indicate the opposite, the observed score variance was due to a
larger amount of random error score variance, leading to a stronger
correction.

@@

Concerning the re-arranged inequality from Appendix A, we identify one
crucial condition described in @eq-ineq_short.

$$R_1^3 \leq R_2$$ {#eq-ineq_short}

As long as $R_1^3$ is smaller or equal compared to $R_2$, @eq-ineq_init
is bound to hold true. This means that as long as the average score
reliability, to the power of 3, does not exceed the reduction in score
variance heterogeneity due to the attenuation correction, @eq-ineq_init
is not violated. For example, assuming the average score reliability is
about .8, as long as $R_2$ is at least .512 or larger, the inequality
holds true and ES heterogeneity grows larger as a consequence of the
attenuation correction procedure. In such cases, the claims made in W&D
and H&S could not hold.

$R_1$ essentially describes the reduction in mean score variance due to
the attenuation correction, and $R_2$ essentially described the
reduction in heterogeneity of score variance due to the attenuation
correction. Therefore, @eq-ineq_init states that for such a correction
procedure to truly reduce ES heterogeneity, the reduction in the
heterogeneity of score variance needs to be substantially larger than
the reduction in its mean values. The mean value reduction carries
stronger weight, as $R_1$ is taken to the power of 3 in @eq-ineq_init.

As long as there are substantial differences in the true score variance
across administrations, it is unlikely that an attenuation correction
can actually produce such a low $R_2$. Generally speaking, $R_1$, and
therefore the reduction in mean score variance, largely depends on the
relative measurement precision. If measurements were made more
precisely, average score reliability and $R_1$ would be higher. On the
other hand $R_2$, and therefore the reduction in heterogeneity of score
variance, depends on how different the measurement precision was at the
individual administrations, but also how different the individual
administrations sites are regarding their true score variance. If the
administration sites carry large differences in true score variance
$R_2$ is unlikely to be substantially smaller than $R_1$.

***\[maybe move to discussion further below\]*** The majority of studies
in the social sciences make use of convenience sampling. Often,
respondents are collected from a local student pool, where students
themselves would choose which studies to partake in. In such cases, we
believe there is little reason for us to expect homogeneity in true
score variance across administration sites. As participants across
administrations self-select into the study, no real sampling process
takes place - implying that we have no reason to assume that the true
score variance is stable across these administration sites. However,
since researchers are well aware of the importance of measurement
precision, which is also held stable more easily, we believe there is
little reason for $R_1$ to be extremely low.
